# Fish Diversity Correlation to Macroinvertebrate Diversity Over Time

## Authors: Andrea Mascioli, Carmel Serban, Nicole Landry, Matthews Do Carmo Nascimiento

Load necessary packages

```{r}
library(tidyverse)
library(here)
library(neonUtilities)
```

Load raw datasets

```{r}
# Fish sampled
water_fishes <- "DP1.20107.001"

# Macroinvertebrates sampled
water_macroinvertebrates <- "DP1.20120.001"

# Temperature Sampled
site_temperatures <- "DP1.00002.001"

# Precipitation Sampled
#site_precipitation <- ""

# Two selected focal sites
my_sites <- c("MAYF", "BIGC")

# Start date (as YYYY-MM)
start_date <- "2018-01"

# End date (as YYYY-MM)
end_date <- "2021-11"

list2env((loadByProduct(dpID = water_fishes, 
                        site = my_sites, 
                        include.provisional = TRUE,
                        startdate = start_date,
                        enddate = end_date)),
         .GlobalEnv)

variables_20107 %>% 
  write_csv(here("data", "raw", "variables_20107.csv"))

fsh_perFish %>%
  write_csv(here("data", "raw", "fsh_perFish.csv"))

list2env((loadByProduct(dpID = water_macroinvertebrates, 
                        site = my_sites, 
                        include.provisional = TRUE,
                        startdate = start_date,
                        enddate = end_date)),
         .GlobalEnv)

variables_20120 %>% 
  write_csv(here("data", "raw", "variables_20120.csv"))

inv_taxonomyProcessed %>% 
  write_csv(here("data", "raw", "inv_taxonomyProcessed.csv"))

list2env((loadByProduct(dpID = site_temperatures, 
                        site = my_sites, 
                        include.provisional = TRUE,
                        startdate = start_date,
                        enddate = end_date)),
         .GlobalEnv)

variables_20120 %>% 
  write_csv(here("data", "raw", "variables_20120.csv"))

inv_taxonomyProcessed %>% 
  write_csv(here("data", "raw", "inv_taxonomyProcessed.csv"))
```

Data wrangling (Selecting fish diversity and macroinvertebrate data)

```{r}
fish <- read.csv(here("data", "raw", "fsh_perFish.csv"))

str(fish)

## Select focal data for fish.

fish_select <- fish %>% 
  select(taxonID, siteID, passStartTime)

## Separate fish diversity by month from other data.

fish_by_month <- fish_select %>% 
  mutate(month = month(passStartTime),
         year = year(passStartTime)) %>% 
  group_by(siteID, year, month) %>% 
  summarize(unique_taxa = n_distinct(taxonID))

print(fish_by_month) 

## write fish_by_month as a separate .csv

write_csv(fish_by_month, here("data", "processed", "fish_by_month.csv"))

# Macroinvertebrate data wrangling:

## Read macroinvertebrate data.

macro_raw <- read_csv(here("data", "raw", "inv_taxonomyProcessed.csv")) 
str(macro_raw)
names(macro_raw)

## Select raw data.

macro_select <- macro_raw %>% 
  select(acceptedTaxonID, collectDate, siteID)

## Separate data to get macroinvertebrate data by month.

 macro_by_month <- macro_select %>% 
  mutate(month = month(collectDate),
         year = year(collectDate)) %>%
  group_by(siteID, year, month) %>% 
  summarize(unique_taxa = n_distinct(acceptedTaxonID))

print(macro_by_month)

```

# Sort by season, not month to get better data feedback
```{r}
# MACROINVERTS

macro_by_season <- macro_select %>%
  mutate(month = month(collectDate),
         year = year(collectDate),
         siteID = factor(siteID),
         season = factor(ifelse(month %in% c(3,4,5), "spring",
                         ifelse(month %in% c(6,7,8), "summer",
                                ifelse(month %in% c(9,10,11), "fall", "winter"))))) %>%
  group_by(siteID, season, year, month) %>%
  summarize(unique_taxa = n_distinct(acceptedTaxonID))
print(macro_by_season)

# FISH

fish_by_season <- fish_select %>%
  mutate(taxonID = factor(taxonID),
         siteID = factor(siteID),
         month = month(passStartTime),
         year = year(passStartTime),
         siteID = factor(siteID),
         season = factor(ifelse(month %in% c(3,4,5), "spring",
                         ifelse(month %in% c(6,7,8), "summer",
                                ifelse(month %in% c(9,10,11), "fall", "winter"))))) %>%
  group_by(siteID, season, year, month) %>%
  summarize(unique_taxa = n_distinct(taxonID),
            total_individuals = n())

print(fish_by_season)

 
```
#Use an ANOVA test to find correlation between unique species and season for each site

```{r}

# MACRO ANOVA

unique_taxa_aov_macro <- aov(unique_taxa ~ season * siteID, data = macro_by_season)

summary(unique_taxa_aov_macro)

# Additive Model
unique_taxa_aov2_macro <- aov(unique_taxa ~ season + siteID, data = macro_by_season)

summary(unique_taxa_aov2_macro)

```
# The data above from the ANOVA test and Additive models show that the siteID's have a significant correlation among them, while season and season:siteID do not have significant correlation. Essentially, the trend we are seeing with seasons and the fluxuations of species diversity are consistant between sites. The p value for siteID was 7.39x10^-7, and the F value was 71.097. 
```{r}
# FISH ANOVA

unique_taxa_aov_fish <- aov(unique_taxa ~ season * siteID, data = fish_by_season)

summary(unique_taxa_aov_fish)

# Additive Model
unique_taxa_aov2_fish <- aov(unique_taxa ~ season + siteID, data = fish_by_season)

summary(unique_taxa_aov2_fish)
```
#The ANOVA test and Additive model reflect the same results as the macroinvertebrate data, with siteID proving significant correlation within them. The p value is 4.37x10^-5 and the F value is 63.984. This data informs us that the fluxuations of species diversity are not only consistant between sites, but also across fish species and macroinvertebrate species, regardless of the site location and geography.


```{r}
# MACRO SITE GRAPH

ggplot(macro_by_season,
       aes(x = season,
           y = unique_taxa,
           fill = siteID)) +
  geom_boxplot() +
  theme_bw()

```
```{r}
# FISH SITE GRAPH

ggplot(fish_by_season,
       aes(x = season,
           y = unique_taxa,
           fill = siteID)) +
  geom_boxplot() +
  facet_wrap(. ~ siteID, 
             scales = "free_y") +
  theme_bw()
```



